Modelling with ANIMO:\\ between Fuzzy Logic and ODE

### 2015-04-29 Paper.tex
Made text and references more consistent. Improved comparison with ODE and fuzzy logic models. Temporarily moved figures inside the text.

\subsection*{A brief introduction to ANIMO}
The user interface of ANIMO is displayed in Figure~\ref{fig:cytoscape}, where we present an example of biological network enriched with dynamic information. ANIMO is based on Cytoscape~\cite{cytoscape}, a software tool developed to represent biological networks. On top of the static topological information displayed in Cytoscape, ANIMO represents biological interactions starting from the basic concept of \emph{activity}: each biological entity in an ANIMO model is considered to be either active or inactive. Activity is to be intended in a very broad sense: for example, an active gene is being transcribed, an active kinase can perform phosphorylations, etc. Each node in an ANIMO network represents both active and inactive entities of the same type, with the relative amount of active entities represented by the node color on a user-configurable scale. Interactions among nodes define how the biological entities in a network influence each other's activity. Only nodes whose activity is larger than 0 can have an influence on their downstream targets, and only if that influence is not counterbalanced by intervening opposite interactions. For example, the interaction $A\rightarrow B$ indicates that node A, if active, will increase the activity of node B. If we add an additional interaction to the example, $C \dashv$\hspace{0.1em}$B$, with C also active, then the activity of B will change depending on the activity levels of A and C, and on the \emph{strength} of their influence. The strength of an ANIMO interaction is a scale factor $k$ which defines the speed at which that interaction occurs: higher values of $k$ give faster interactions. These $k$-values are the only parameters needed in an ANIMO model, and can be given in either a quantitative (as real numbers) or a qualitative way, choosing among self-explanatory descriptions such as ``very slow'', ``slow'', ``medium'', ``fast'', ``very fast''.

@@

The output of the ANIMO model in Figure~\ref{fig:drosophila-model} closely matches the original ODE model.
In particular, the oscillations in both models show the same periods and phases (see Suppl. Fig~\ref{suppl-fig:grafici-drosophila}).
% Due to the compositional nature of \tas\, ANIMO allows for intuitive \emph{in silico} knock-out experiments,
% by right-clicking a node in the model and disabling it. Such experiments have been done
% before~\cite{drosophila-ode-model} and give similar results in our model.
Details of the comparison
between the ANIMO model and the original ODE model is given in Supplementary Section~\ref{suppl-sec:animo-drosophila}.
% \\
% {\Large [[Insert description of K.O. experiments with Drosophila model]]}

\subsection*{Using ANIMO to generate hypotheses}\label{subsec:case-study-larger}
We now present a comparison with an existing fuzzy logic model, which we use also as an example of how ANIMO can be used to create reference networks and help to reason on complex biological events.

We constructed a model of the signaling network downstream of TNF$\alpha$ and EGF, formalizing
the crosstalk that takes place between the pathways at different levels of cellular regulation.
We first modeled the two pathways in isolation~(Suppl. Figs.~\ref{fig:large-model-tnf}, \ref{fig:large-model-egf}),
using information on protein interactions from
the KEGG~\cite{kegg} and phosphosite~\cite{phosphosite} databases. These models were fitted to experimental data
from previous studies~\cite{pathway-compendium,pathway-autocrine}.
We then merged the two pathways into a single model and added autocrine crosstalk between the pathways that has been
described in~\cite{pathway-autocrine}.
Briefly, stimulation with TNF$\alpha$ ({\sf TNFa} in the model) leads to a rapid release of TGF$\alpha$ ({\sf TGFa}),
which activates the EGF receptor ({\sf EGFR}).
This activation causes secretion of IL-1$\alpha$ ({\sf IL-1a}) at later time points.
The effect of IL-1$\alpha$ is down-regulated by the secretion of IL-1 receptor antagonist ({\sf IL-1ra})
downstream of TNF$\alpha$.
The resulting model (Fig.~\ref{fig:large-model-no-hypotheses}) was compared to the experimental data
for treatments with 100 ng/ml TNF alone and 100 ng/ml EGF alone (data not shown)~\cite{pathway-compendium}.

At this point, the behavior of the model deviated from the data for some of the nodes.
Changing the parameters of the model was not enough to reproduce the behaviour shown by experimental
data. This is an interesting situation, as it requires changes in the topology of
the model, which can be interpreted as new hypotheses. Below, we give two examples and show how
adaptation of the model can be used to generate novel testable hypotheses.

Experimentally,
treatment with TGF$\alpha$ alone does not lead to secretion of IL-1$\alpha$. Instead, a co-stimulation with
TGF$\alpha$ and TNF$\alpha$ is required~\cite{pathway-autocrine}.
However, in the first version of the model, treatment with TGF$\alpha$ was sufficient for
IL-1$\alpha$ expression (Fig.~\ref{fig:large-model-graph1}). Given the time delay until secretion of IL-1$\alpha$, it can be
expected that \emph{de novo} synthesis of IL-1$\alpha$ is required and that both
TNF$\alpha$ and TGF$\alpha$ are needed to activate transcription of the IL-1$\alpha$ gene.
JNK1 and ERK signal downstream of TNF$\alpha$ and TGF$\alpha$, respectively, and are known
to affect the activity of multiple transcription factors. We altered the model to make
activation of IL-1$\alpha$ expression dependent on both JNK1 activity and ERK activity
(Suppl. Fig.~\ref{fig:large-model-hypotheses}, arrows linking {\sf JNK1} and {\sf ERK} to {\sf IL-1a gene}).
After this modification to the model, IL-1$\alpha$ was no longer secreted
upon stimulation with TGF$\alpha$ alone, which greatly improved the fit between the measured IL-1$\alpha$
levels and the model (Fig.~\ref{fig:large-model-graph2}). This hypothesis could now be used to
design a new experiment to validate IL-1$\alpha$ as a target of combined JNK1 activity and ERK activity in
HT-29 cells. For example, kinase inhibitors specific to JNK1 and ERK could be used to confirm that activity of
both kinases is required for expression and secretion of IL-1$\alpha$. Performing the experiment is beyond
the scope of this study, but this hypothesis finds support in literature.
Transcription factors c-Jun and c-Fos together
form a heterodimer known as AP-1 and are activated by JNK1 and ERK,
respectively~\cite{jnk-signaling,cfos-cjun}. AP-1 has been reported to bind to the
promoter of IL-1$\alpha$, providing evidence for a role in the regulation of IL-1$\alpha$
expression~\cite{ap1-il1a}. Based on these findings in literature we included c-Jun and
c-Fos in our model as transcriptional activators of IL-1$\alpha$ (Fig.\ref{fig:large-model-complete}).

As a second example, we considered the behaviour of JNK1 and MK2. In the model, both
proteins were located downstream of TNF$\alpha$ but not TGF$\alpha$ or EGF. Hence, the
model did not show an effect of C225, a pharmacological inhibitor of ligand-EGFR
binding, on activation of JNK1 or MK2 after stimulation with TNF$\alpha$. However, experimental
data show that C225 strongly reduces activation of JNK1 and MK2 upon stimulation with
TNF$\alpha$~\cite{pathway-autocrine}.
This fact is indicative of a role for EGFR in activation of JNK1 and MK2. Since both JNK1 and MK2
are located downstream of MEKK1, we hypothesized that activation
of MEKK1 is dependent on
both TNF$\alpha$-signalling and TGF$\alpha$-signalling. In the model we added a new
hypothetical node {\sf Hyp~2} (hypothesis~2) to link EGFR to MEKK1 (Suppl. Fig.~\ref{fig:large-model-hypotheses}).
This addition led to an improved fit of the model to the data upon treatment with TNF$\alpha$ + C225:
activation of both MK2 and JNK1 was strongly suppressed by C225 (Fig.~\ref{fig:large-model-graph4}).
Stimulation with EGF alone did not lead to activation of JNK1 and MK2.
These data support the validity of the modification to the model.
Further support for a link between EGFR and MEKK1 was found in literature. Specifically,
Ras has been reported as a direct activator of
MEKK1~\cite{ras-mekk1}. EGFR is a well-known and potent activator of Ras,
which is why it was already in our network~\cite{kegg}.
Other studies also report activation of JNK1 and phosphorylation of c-Jun downstream of Ras, which is consistent with
an interaction between Ras and MEKK1~\cite{cfos-cjun,ras-jnk1}.
Based on these findings, we adapted
our model by removing the {\sf Hyp~2} node and creating a direct interaction between Ras
and MEKK1 (Fig.~\ref{fig:large-model-complete}). Experimentally, the role of Ras could be confirmed by using a
pharmacological inhibitor of Ras activity, and measuring the effect of this inhibitor on the activation of JNK1 and MK2.
Together, our model suggests that EGFR activity is required
but not sufficient for activation of JNK1 and MK2 in HT-29 cells.

There are other nodes for which the experimental data deviates from the model in one or more of the experimental conditions.
A comparison between model and experimental data can be found in Figures~\ref{fig:differences1}, \ref{fig:differences2} and~\ref{fig:differences3}.
A complete deciphering of the signalling events
in this biological system is outside the scope of this paper. Instead, we illustrated how interactive modelling of the dynamic behaviour
of a signal transduction network can be used to extend previous pathway topologies and can lead to the generation of novel hypotheses.

\section*{Discussion}
\subsection*{Final remarks on the models}
In the Results section, we described the construction of an ANIMO
model of the circadian clock in \emph{Drosophila Melanogaster}. This model
captured the dynamics of the regulatory network and led to similar
conclusions as an ODE model that had been
published previously~\cite{drosophila-ode-model}. This finding supports the use of
the series of modeling abstractions on which ANIMO is based. The biggest
difference between the construction of these models is that the model in~\cite{drosophila-ode-model}
is constructed by writing a system of mathematical equations, together
with an algorithm for simulation. In ANIMO, instead, a number of network
nodes is drawn for the molecules involved.
These nodes are then linked by directed
interactions that represent cause-and-effect relationships, with a single parameter
that defines the strength of each
interaction. This is a more intuitive approach to construct a model.
Further contributing to an interactive modeling process
is the compositionality of the model. Each node in the network
can be disabled at any time by the user, or extra nodes can be added,
without having to change any of the existing interactions.

In the Results section we also showed the construction of an executable model
of signaling events downstream of
TNF$\alpha$ and EGF in human colon carcinoma cells. This data set has been used for
previous modeling studies, based on partial least-squares regression and fuzzy
logic~\cite{pathway-leastsquare,pathway-fuzzy}.
The partial least-squares regression model describes an abstract data-driven model
that uses statistical correlations
to relate signal transduction events to various cellular decisions. This type of modeling is
very useful in uncovering new and unexpected relations. It is also successful in making
predictions, but gives little direct in the dynamic behavior of the network. Fuzzy
logic analysis led to a model that gave a better fit to the dynamic network behavior than
discrete logic (Boolean) models. Inspection of the inputs to the logical gates that were used
to model protein behavior led to the prediction of novel interactions between proteins,
showing the usefulness of this approach. For most of the proteins, such as JNK1, time was
used as an input parameter. For example, the logical gates ``if TNF$\alpha$ is high
\emph{AND} time is low, then JNK1 is high'' and ``if TNF$\alpha$ is high \emph{AND} time is
high, then JNK1 is low'' were used to
describe the dynamic behavior of JNK1. Although this leads to a representative
description of the dynamic behavior of JNK1, peaks in protein activity at early time points were
not reproduced by the fuzzy logic model. Moreover, it gives no insight in the molecular interactions
that are involved in activation or inhibition.

@@

Going beyond the user interface, there are a number of ``pros and cons'' for using ANIMO and \tas\
in the biological context. First and foremost, being \tas\ an executable formal language,
a state space can be derived from a \tas\ model. This means that state space-related analyses
such as model checking can be performed on \tas: as explained in the Results section, this can
be done directly in ANIMO, bypassing the need for a deeper understanding of \tas\ as
ANIMO acts as an intermediary towards the powerful model checking tool UPPAAL.

@@

All reactants modeled in ANIMO are assumed to be present at the same
(unitary) concentration. While this assumption is not always applicable, it encourages
abstract thought: many biological processes can be represented as networks driven by
\emph{activity-based} interactions. Even if with a limited scope, ANIMO can be applied
also in the analysis of metabolic processes, considering the production of protein $A$ as
the activation of node $A$, with degradation being represented as inhibition. As stated before,
it cannot be expected from such models to be a completely realistic representation of their target biological
processes, but they can still be an useful tool. This can be seen for example in the Drosophila
model in the Results section, where mRNA and protein concentrations were abstracted to
activity-driven processes in the ANIMO model.

@@

In this paper we discussed the placement of ANIMO among other modeling paradigms and tools, highlighting ANIMO's strong points. Our aim is not to promote the use of ANIMO in any modeling effort, but to show where it best shines.

@@

Timed Automata have been shown to be a powerful formalism to model biological processes
~\cite{ta-siebert,bartocci-oscillators,oded-ode-ta-discretization}. A timed automaton consists of locations
and transitions between these locations (see Fig.~\ref{fig:abstraction-mek-erk}), and a system of timed automata can be
used to model a system of interacting molecules. At any time, each automaton is in a specific location, and together
these locations represent the current state of the biological system. Each timed automaton can have one or more local clocks
associated to it, allowing temporal control of transitions between locations. These transitions are used to
represent interactions between molecules. Fast interactions take less time than slow interactions
to perform an activation or inhibition step. We have previously described in detail how the
scenarios presented in Section~\ref{subsec:abstractions} can be used to calculate the timing of molecular
interactions to give a description of network dynamics~\cite{animo-ieee}. Figure~\ref{fig:abstraction-mek-erk}
presents a small example that illustrates the basic properties of \tas.
This model describes the activation of ERK by MEK\footnote{All acronyms used in this paper
and their corresponding UniProt IDs are listed in Suppl. Sect.~\ref{suppl-sec:names}.}.

\subsection*{Example: building a model based on data}\label{subsec:case-study}
To illustrate the use of ANIMO in a practical environment, we will demonstrate the generation of a basic
version of the model described in Section~\ref{subsec:case-study-larger}. The model is based on a literature compendium of
signal transduction events in HT-29 human colon carcinoma cells~\cite{pathway-compendium}. This data set comprises triplicate
measurements of 11 different protein activities or post-translational modification states at 13 time points after
treatment with different combinations of tumor necrosis factor-$\alpha$ (TNF$\alpha$), epidermal growth factor (EGF) and insulin.
The data set contains relative protein levels and activities, and no absolute quantities, which is the typical situation in biochemistry.
To start, we normalized measurements for each protein to the
maximum value in the complete experiment, resulting in a nondimensionalized data set that is suitable for use with ANIMO.

In Figure~\ref{fig:small-model}, we show the stepwise construction of a model of a small part of the network that is
able to account for measured variations in activity of inhibitor of nuclear factor kappa-B kinase (IKK), c-Jun N-terminal kinase-1 (JNK1),
mitogen-activated protein kinase-activated protein kinase 2 (MK2), Caspase 8 (casp-8) and Caspase 3 (casp-3) upon stimulation with 100
ng/ml TNF$\alpha$. In this example we aimed for inclusion of a minimum number of nodes in the network, while preserving biological relationships.
Multistep cascades were aggregated into a single step when possible. Parameters for all reactions were set manually, resulting in a close
match between the model and the patterns observed in the dataset.

>>>>

\subsection*{A brief introduction to ANIMO}
The user interface of ANIMO is displayed in Figure~\ref{fig:cytoscape},
where we present an example of biological network enriched with dynamic information.
ANIMO is based on Cytoscape~\cite{cytoscape}, a software tool developed to represent biological networks.
On top of the static topological information displayed in Cytoscape, ANIMO represents
biological interactions starting from the basic concept of \emph{activity}:
each biological entity in an ANIMO model is considered to be either active or inactive.
Activity is to be intended in a very broad sense: for example, an active gene is being transcribed,
an active kinase can perform phosphorylations, etc. Each node in an ANIMO network represents
both active and inactive entities of the same type, with the relative amount of active entities
represented by the node color on a user-configurable scale. Interactions among nodes define
how the biological entities in a network influence each other's activity.
Only nodes whose activity is larger than 0 can have an 
influence on their downstream targets, and only if that influence is not counterbalanced by intervening opposite interactions.
For example, the interaction $A\rightarrow B$ (read ``A activates B'') indicates that node A, if active,
will increase the activity of node B. If we add an additional interaction to the example,
$C \dashv$\hspace{0.1em}$B$ (``C inhibits B''), with C also active, then the activity of B will change depending on the
activity levels of A and C, and on the \emph{strength} of their influence.
The strength of an ANIMO interaction is a scale factor $k$ which defines the speed at which
that interaction occurs: higher values of $k$ give faster interactions. These $k$-values are
the only parameters needed in an ANIMO model, and can be given in either
a quantitative (as real numbers) or a qualitative way, choosing among self-explanatory descriptions
such as ``very slow'', ``slow'', ``medium'', ``fast'', ``very fast''.

\begin{figure*}[htbp]
\begin{center}
   \includegraphics[width=.85\textwidth]{images/screenshotTNFmodelSmall}
\end{center}
\caption{\csentence{The Cytoscape user interface running the ANIMO plug-in.}
The \emph{Network} panel in the center contains the nodes-edges
model of the example TNF$\alpha$ patwhay (see Methods Section), with
colors indicating node activity levels and shapes representing different protein categories (see the \emph{Legend} on the left).
The \emph{Results Panel} on the right contains a graph plotting activity levels of selected nodes
during the first 24 hours of evolution of the model. The slider under the graph
allows the user to select the time instant (marked as a vertical red line in the graph) on which
the colors nodes in the \emph{Network} are based.
The series with the {\sf \_data} suffix is experimental
data from~\cite{pathway-compendium}, considering a treatment with 100~ng/ml TNF$\alpha$.\\
All acronyms used in this paper
and their corresponding UniProt IDs are listed in Suppl. Sect.~\ref{suppl-sec:names}.
\label{fig:cytoscape}}
\end{figure*}

@@

\def\drosophilaGraphScale{0.069}%0.123}
\begin{figure*}[!htpb]
\begin{center}
\includegraphics[width=0.75\textwidth]{images/drosophila_model5}%0.85\textwidth
% \subfloat[\label{subfig:drosophila-mrna}]{\includegraphics[scale=\drosophilaGraphScale]{images/drosophila_graph2}}\
% \subfloat[\label{subfig:drosophila-proteins}]{\includegraphics[scale=\drosophilaGraphScale]{images/drosophila_graph1}}
\end{center}
\caption{\csentence{ANIMO model of the circadian clock in \emph{Drosophila Melanogaster}.}
Auto-regulatory negative feedback loops are present on each of the nodes of
the network, following the original model in~\cite{drosophila-ode-model}. These feedback loops ensure that
protein levels decrease over time when activating inputs are absent. The feedback loops are not represented here
for cosmetic reasons only.
% \\ ANIMO plots of the concentration of
% mRNA~{\bfseries \protect\subref{subfig:drosophila-mrna}}
% and proteins~{\bfseries \protect\subref{subfig:drosophila-proteins}} over a period of 48~hours.\\
Naming conventions follow the same rules
as in the original model, with lower-case names representing mRNA, and upper-case names indicating proteins.
}\label{fig:drosophila-model}
\end{figure*}

The output of the ANIMO model in Figure~\ref{fig:drosophila-model} matches the original ODE model.
In particular, the oscillations in both models show the same periods and phases (see Suppl. Fig~\ref{suppl-fig:grafici-drosophila}).
% Due to the compositional nature of \tas\, ANIMO allows for intuitive \emph{in silico} knock-out experiments,
% by right-clicking a node in the model and disabling it. Such experiments have been done
% before~\cite{drosophila-ode-model} and give similar results in our model.
Details of the comparison
between the ANIMO model and the original ODE model is given in Supplementary Section~\ref{suppl-sec:animo-drosophila}.
% \\
% {\Large [[Insert description of K.O. experiments with Drosophila model]]}

\subsection*{Using ANIMO to generate hypotheses}\label{subsec:case-study-larger}
We now present a comparison with an existing fuzzy logic model, which we use also as an example of how
ANIMO can be used to create reference models and help to reason on complex biological events.

We constructed a model of the signaling network downstream of TNF$\alpha$ and EGF
in HT-29 human colo carcinoma cells, formalizing
the crosstalk that takes place between the pathways at different levels of cellular regulation.
We first modeled the two pathways in isolation~(Suppl. Figs.~\ref{fig:large-model-tnf}, \ref{fig:large-model-egf}),
using information on protein interactions from
the KEGG~\cite{kegg} and phosphosite~\cite{phosphosite} databases. These models were fitted to experimental data
from previous studies~\cite{pathway-compendium,pathway-autocrine}.
We then merged the two pathways into a single model and added autocrine crosstalk between the pathways that has been
described in~\cite{pathway-autocrine}.
Briefly, stimulation with TNF$\alpha$ ({\sf TNFa} in the model) leads to a rapid release of TGF$\alpha$ ({\sf TGFa}),
which activates the EGF receptor ({\sf EGFR}).
This activation causes secretion of IL-1$\alpha$ ({\sf IL-1a}) at later time points.
The effect of IL-1$\alpha$ is down-regulated by the secretion of IL-1 receptor antagonist ({\sf IL-1ra})
downstream of TNF$\alpha$.
The resulting model (Fig.~\ref{fig:large-model-all}A) was compared to the experimental data
for treatments with 100 ng/ml TNF alone and 100 ng/ml EGF alone (data not shown)~\cite{pathway-compendium}.

At this point, the behavior of the model deviated from the data for some of the nodes.
Changing the parameters of the model was not enough to reproduce the behaviour shown by experimental
data. This is an interesting situation, as it requires changes in the topology of
the model~\cite{animo-syncop}, which can be interpreted as new hypotheses. Below, we give two examples and show how
adaptation of the model can be used to generate novel testable hypotheses.

Experimentally,
treatment with TGF$\alpha$ alone does not lead to secretion of IL-1$\alpha$. Instead, a co-stimulation with
TGF$\alpha$ and TNF$\alpha$ is required~\cite{pathway-autocrine}.
However, in the first version of the model, treatment with TGF$\alpha$ was sufficient for
IL-1$\alpha$ expression (Fig.~\ref{fig:large-model-all}B). Given the time delay until secretion of IL-1$\alpha$, it can be
expected that \emph{de novo} synthesis of IL-1$\alpha$ is required and that both
TNF$\alpha$ and TGF$\alpha$ are needed to activate transcription of the IL-1$\alpha$ gene.
JNK1 and ERK signal downstream of TNF$\alpha$ and TGF$\alpha$, respectively, and are known
to affect the activity of multiple transcription factors. We altered the model to make
activation of IL-1$\alpha$ expression dependent on both JNK1 activity and ERK activity
(Suppl. Fig.~\ref{fig:large-model-hypotheses}, arrows linking {\sf JNK1} and {\sf ERK} to {\sf IL-1a gene}).
After this modification to the model, IL-1$\alpha$ was no longer secreted
upon stimulation with TGF$\alpha$ alone, which greatly improved the fit between the measured IL-1$\alpha$
levels and the model (Fig.~\ref{fig:large-model-complete}B). This hypothesis could now be used to
design a new experiment to validate IL-1$\alpha$ as a target of combined JNK1 activity and ERK activity in
HT-29 cells. For example, kinase inhibitors specific to JNK1 and ERK could be used to confirm that activity of
both kinases is required for expression and secretion of IL-1$\alpha$. Performing the experiment is beyond
the scope of this study, but this hypothesis finds support in literature.
Transcription factors c-Jun and c-Fos together
form a heterodimer known as AP-1 and are activated by JNK1 and ERK,
respectively~\cite{jnk-signaling,cfos-cjun}. AP-1 has been reported to bind to the
promoter of IL-1$\alpha$, providing evidence for a role in the regulation of IL-1$\alpha$
expression~\cite{ap1-il1a}. Based on these findings in literature we included c-Jun and
c-Fos in our model as transcriptional activators of IL-1$\alpha$ (Fig.\ref{fig:large-model-complete}A).

\begin{figure*}[!htpb]
\begin{center}
% \subfloat[\label{fig:large-model-complete}]{\includegraphics[scale=\largeModelScale]{images/large_network_legendg}}\\
%  \subfloat[\label{fig:large-model-graph2}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_vs_TNF100_hyp1_IL-1a4}}
%  \subfloat[\label{fig:large-model-graph4}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_hyp2_JNK1_MK2}}
  \includegraphics[width=0.9\textwidth]{images/large_network_legendg_oneImage}
\end{center}
\caption{\csentence{Signaling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells: updated version.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-complete}}
The model for the merged TNF$\alpha$ and EGF pathways
after addition of the two hypotheses (highlighted).
Hypothesis 1 assumes IL-1$\alpha$ expression to depend on AP-1 activity, which in turn requires
both c-Jun en c-Fos to be activated by JNK1 and ERK, respectively. Hypothesis 2 assumes RAS as an activator
of MEKK1. Node colors represent the activity levels $15$ minutes
after stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph2}}~
After the addition of the first hypothesis (activation of IL-1$\alpha$ production depending both
on JNK1 and ERK): production of IL-1$\alpha$ after stimulation with 100 ng/ml TNF$\alpha$ (series {\sf IL-1a~(TNFa)})
compared with stimulation with 100 ng/ml TGF$\alpha$ (series {\sf IL-1a~(TGFa)}) (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph4}}~
After the addition of the second hypothesis (activation of MEKK1 downstream of EGFR):
stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
Suppl. Sect.~\ref{suppl:parameters-tnf-egf} explains how the dosage of 5 ng/ml TNF$\alpha$ was represented in the model.\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-complete}
\end{figure*}

As a second example, we considered the behaviour of JNK1 and MK2. In the model, both
proteins were located downstream of TNF$\alpha$ but not TGF$\alpha$ or EGF. Hence, the
model did not show an effect of C225, a pharmacological inhibitor of ligand-EGFR
binding, on activation of JNK1 or MK2 after stimulation with TNF$\alpha$. However, experimental
data show that C225 strongly reduces activation of JNK1 and MK2 upon stimulation with
TNF$\alpha$~\cite{pathway-autocrine}.
This fact is indicative of a role for EGFR in activation of JNK1 and MK2. Since both JNK1 and MK2
are located downstream of MEKK1, we hypothesized that activation
of MEKK1 is dependent on
both TNF$\alpha$-signalling and TGF$\alpha$-signalling. In the model we added a new
hypothetical node {\sf Hyp~2} (hypothesis~2) to link EGFR to MEKK1 (Suppl. Fig.~\ref{fig:large-model-hypotheses}).
This addition led to an improved fit of the model to the data upon treatment with TNF$\alpha$ + C225:
activation of both MK2 and JNK1 was strongly suppressed by C225 (Fig.~\ref{fig:large-model-complete}C).
Stimulation with EGF alone did not lead to activation of JNK1 and MK2.
These data support the validity of the modification to the model.
Further support for a link between EGFR and MEKK1 was found in literature. Specifically,
Ras has been reported as a direct activator of
MEKK1~\cite{ras-mekk1}. EGFR is a well-known and potent activator of Ras,
which is why it was already in our network~\cite{kegg}.
Other studies also report activation of JNK1 and phosphorylation of c-Jun downstream of Ras, which is consistent with
an interaction between Ras and MEKK1~\cite{cfos-cjun,ras-jnk1}.
Based on these findings, we adapted
our model by removing the {\sf Hyp~2} node and creating a direct interaction between Ras
and MEKK1 (Fig.~\ref{fig:large-model-complete}A). Experimentally, the role of Ras could be confirmed by using a
pharmacological inhibitor of Ras activity, and measuring the effect of this inhibitor on the activation of JNK1 and MK2.
Together, our model suggests that EGFR activity is required
but not sufficient for activation of JNK1 and MK2 in HT-29 cells.

There are other nodes for which the experimental data deviates from the model in one or more of the experimental conditions.
A comparison between model and experimental data can be found in Figures~\ref{fig:differences1}, \ref{fig:differences2} and~\ref{fig:differences3}.
Comparing these results with the ones from~\cite{pathway-fuzzy} shows a better fit of the ANIMO model, which is also intrinsically
more precise, being more mechanistic in nature (see Fig.~\ref{fig:animo-spectrum}).
A complete deciphering of the signalling events
in this biological system is outside the scope of this paper. Instead, we illustrated how interactive modelling of the dynamic behaviour
of a signal transduction network can be used to extend previous pathway topologies and can lead to the generation of novel hypotheses.

\section*{Discussion}
\subsection*{Final remarks on the models}
In the Results section, we described the construction of an ANIMO
model of the circadian clock in \emph{Drosophila Melanogaster}. This model
captured the dynamics of the regulatory network and led to similar
conclusions as an ODE model that had been
published previously~\cite{drosophila-ode-model}. This finding supports the use of
the series of modeling abstractions on which ANIMO is based. The biggest
difference between the construction of these models is that the model in~\cite{drosophila-ode-model}
is constructed by writing a system of mathematical equations, together
with an algorithm for simulation. In ANIMO, instead, a number of network
nodes is drawn for the molecules involved.
These nodes are then linked by directed
interactions that represent cause-and-effect relationships, with a single parameter
that defines the strength of each
interaction. This is a more intuitive approach to construct a model.
Further contributing to an interactive modeling process
is the compositionality of the model. Each node in the network
can be disabled at any time by the user, or extra nodes can be added,
without having to change any of the existing interactions.
All this comes at the price of lower model precision: the curves representing
oscillation of protein activities in the ANIMO model are not as smooth
nor as precise as those obtained from the original ODE model (see Suppl. Fig.~\ref{suppl-fig:grafici-drosophila}).

In the Results section we also showed the construction of an executable model
of signaling events downstream of
TNF$\alpha$ and EGF in human colon carcinoma cells. This data set has been used for
previous modeling studies, based on partial least-squares regression and fuzzy
logic~\cite{pathway-leastsquare,pathway-fuzzy}.
The partial least-squares regression model describes an abstract data-driven model
that uses statistical correlations
to relate signal transduction events to various cellular decisions. This type of modeling is
very useful in uncovering new and unexpected relations. It is also successful in making
predictions, but gives little direct insight in the dynamic behavior of the network (see Fig.~\ref{fig:animo-spectrum}). Fuzzy
logic analysis led to a model that gave a better fit to the dynamic network behavior than
discrete logic (Boolean) models. Inspection of the inputs to the logical gates that were used
to model protein behavior led to the prediction of novel interactions between proteins,
showing the usefulness of this approach. For most of the proteins, such as JNK1, time was
used as an input parameter. For example, the logical gates ``if TNF$\alpha$ is high
\emph{AND} time is low, then JNK1 is high'' and ``if TNF$\alpha$ is high \emph{AND} time is
high, then JNK1 is low'' were used to
describe the dynamic behavior of JNK1. Although this leads to a representative
description of the dynamic behavior of JNK1, peaks in protein activity at early time points were
not reproduced by the fuzzy logic model. Moreover, it gives no insight in the molecular interactions
that are involved in activation or inhibition.

@@

Going beyond the user interface, there are a number of ``pros and cons'' for using ANIMO and \tas\
in the biological context. First and foremost, being \tas\ an executable formal language,
a state space can be derived from a \tas\ model. This means that state space-related analyses
such as model checking can be performed on \tas: this can
be done directly in ANIMO, bypassing the need for a deeper understanding of \tas\ as
ANIMO acts as an intermediary towards the powerful model checking tool UPPAAL.

@@

All reactants modeled in ANIMO are assumed to be present at the same
(unitary) concentration. While this assumption is not always applicable, it encourages
abstract thought: many biological processes can be represented as networks driven by
\emph{activity-based} interactions (see Methods Section). Even if with a limited scope, ANIMO can be applied
also in the analysis of metabolic processes, considering the production of protein $A$ as
the activation of node $A$, with degradation being represented as inhibition. As stated before,
it cannot be expected from such models to be a completely realistic representation of their target biological
processes, but they can still be an useful tool. This can be seen for example in the Drosophila
model in the Results section, where mRNA and protein concentrations were abstracted to
activity-driven processes in the ANIMO model.

@@

In this paper we discussed the placement of ANIMO among other modeling paradigms and tools,
highlighting ANIMO's strong points. The aim is not to promote the use of ANIMO in any modeling effort, but to show where it best shines.

@@

Timed Automata have been shown to be a powerful formalism to model biological processes
~\cite{ta-siebert,bartocci-oscillators,oded-ode-ta-discretization}. A timed automaton consists of locations
and transitions between these locations (see Fig.~\ref{fig:abstraction-mek-erk}), and a system of timed automata can be
used to model a system of interacting molecules. At any time, each automaton is in a specific location, and together
these locations represent the current state of the biological system. Each timed automaton can have one or more local clocks
associated to it, allowing temporal control of transitions between locations. These transitions are used to
represent interactions between molecules. Fast interactions take less time than slow interactions
to perform an activation or inhibition step. We have previously described in detail how approximated
scenarios~\cite{animo-ieee} (see Section~\ref{subsec:abstractions}) can be used to calculate the timing of molecular
interactions to give a description of network dynamics. Figure~\ref{fig:abstraction-mek-erk}
presents a small example that illustrates the basic properties of \tas.
This model describes the activation of ERK by MEK.

\subsection*{Example: building a model based on data}\label{subsec:case-study}
To illustrate the use of ANIMO in a practical environment, we will demonstrate the generation of a basic
version of the model described in the Results Section. The model is based on a literature compendium of
signal transduction events in HT-29 human colon carcinoma cells~\cite{pathway-compendium}. This data set comprises triplicate
measurements of 11 different protein activities or post-translational modification states at 13 time points after
treatment with different combinations of tumor necrosis factor-$\alpha$ (TNF$\alpha$), epidermal growth factor (EGF) and insulin.
The data set contains relative protein levels and activities, and no absolute quantities, which is the typical situation in biochemistry.
To start, we normalized measurements for each protein to the
maximum value in the complete experiment, resulting in a nondimensionalized data set that is suitable for use with ANIMO (see Suppl. Sect.~\ref{suppl-sec:normalization}).
